%*******************************************************
% Acknowledgments
%*******************************************************
\pdfbookmark[0]{Acknowledgments}{acknowledgments}
\addcontentsline{toc}{chapter}{Acknowledgments}

\begin{flushright}{\slshape    
    Gratitude is the recognition that nothing stands alone— \\
    that every thought is a collaboration, every insight a gift, \\
    every achievement a trace of voices heard and unheard.} \\ \medskip
    --- José Aram Méndez Gómez
\end{flushright}

\bigskip

\begingroup
\let\clearpage\relax
\let\cleardoublepage\relax
\chapter*{Acknowledgments}

I would like to express my deepest gratitude to my company supervisor, 
\textbf{Alexander-A Efremov} of \textbf{Deutsche Bank AG}. His guidance was 
not only instrumental but foundational to the direction and realization of 
this thesis. His technical insight, critical feedback, and continuous support 
shaped both the research approach and its practical execution in decisive ways.

I also wish to thank my academic supervisor, \textbf{Prof. Dr. Markus Esch} 
of \textbf{Hochschule für Technik und Wirtschaft des Saarlandes} for his constructive 
feedback and encouragement, which ensured the academic rigor of this work.

This project was truly a collaborative effort across multiple teams. I extend my 
sincere appreciation to the \textbf{distinct engineers} from the various teams at 
\textbf{Deutsche Bank AG} who formed our cross-functional working group. Each 
brought unique expertise—from payment processing and validation logic to infrastructure 
and user experience—that was essential to this system's development. Your willingness 
to share knowledge across team boundaries, provide critical feedback during design 
reviews, and collaborate on integrating this solution into the broader eBridge-EU 
ecosystem exemplified the best of engineering collaboration. This thesis would not 
have been possible without your collective technical excellence and collegial support.

My thanks extend as well to \textbf{Deutsche Bank AG} for enabling this thesis 
within an innovative and challenging professional environment. The opportunity 
to work at the intersection of academic research and real-world application 
has been invaluable.

\endgroup
